Solution to ``Chapter 2: Basic tail and concentration bounds"

### 2020-07-31 skills/math/tail_bound/tailbound_solution.tex
tailbound

@@

	To evaluate $I_1$, we spilt the integral to avoid the minimization. Solving $1 = C \exp \of{-\frac{t^2}{4 \nu^2}} $, the minimization term becomes
	\begin{align}
		\min \offf{1, C \exp \of{-\frac{t^2}{4 \nu^2}}} = \begin{cases}
			\quad 1 & \text{when } t \leq 2\nu \sqrt{\log C},\\
			\quad C \exp \of{-\frac{t^2}{4 \nu^2}} & \text{when } t \geq 2\nu \sqrt{\log C}.
		\end{cases}
	\end{align}
	Therefore, we evaluate $I_1$ as follows,
	\begin{align}
		I_1 &= \int_{0}^{ 2\nu \sqrt{\log C}} 1 dt + \int_{2\nu \sqrt{\log C}}^{+\infty} C \exp \of{-\frac{t^2}{4 \nu^2}} dt\\
		(\text{let}\quad y = \frac{t}{2\nu} - \sqrt{\log C})\quad &=  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2 - 2y \sqrt{\log C}} dy\\
		&\leq  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2} dy.
	\end{align}
	By Gaussian integral $\int_{0}^{+\infty} e^{-x^2} dx = \frac{\sqrt{\pi}}{2}$, we obtain $I_1 \leq 2\nu (\sqrt{\pi} + \sqrt{\log C})$. Similarly, we evaluate $I_2$ as follows 
	\begin{align}
		I_2 &= \int_{0}^{4B \log C} 1 dt + \int_{4B \log C}^{+\infty} C \exp \of{-\frac{t}{4B}} dt\\
		 &= 4B (\log C + 1).
	\end{align}
	Hence, we obtain the expectation of $Z$,
	\begin{equation}
		\bbE[Z] = I_1 + I_2 \leq  2\nu (\sqrt{\pi} + \sqrt{\log C}) + 4B (\log C + 1).
	\end{equation}

@@

\begin{rmk}
	For all nonnegative random variables satisfying the Bernstein-type inequality \eqref{eq:berncond1} with parameters $(\nu,B,C)$, the expectations of the variables are upper bounded by a function of $(\nu,B,C)$. Particularly, for a zero-mean random variable $X$ satisfying the Bernstein condition with parameter $b$, $|X|$ satisfies the inequality \eqref{eq:berncond1} with parameters $(\sigma,b,2)$, where $\sigma^2 = \var(X)$.  The expectation of the absolute variable $|X|$ depends on the variance $\sigma^2$ and the parameter $b$.
\end{rmk}

\subsection{Exercise 2.9}
\begin{lem}
	Let $\{X_i\}_{i=1}^n$ be a sequence of i.i.d.\ Bernoulli variables with parameter $\alpha \in (0,1/2]$. Consider the binomial random variable $Z_n = \sum_{i=1}^n X_i$. The tail probability of $Z_n$ is upper bounded as
	\begin{equation}\label{eq:lembinupper}
		\bbP\off{Z_n \leq \delta n} \leq e^{-n D\of{\delta \Vert \alpha}}, \quad \text{for all }\delta \in (0,\alpha),
	\end{equation}
	where the quantity 
	\begin{equation}
		D(\delta \Vert \alpha) \colonequals \delta \log \frac{\delta}{\alpha} + (1-\delta) \log \frac{1-\delta}{1-\alpha}
	\end{equation}
	is the KL divergence between the Bernoulli distributions with parameters $\delta$ and $\alpha$, respectively.

@@

\begin{proof}
    Applying the Chernoff inequality to the random variable $Z_n$, we have
    \begin{equation}\label{eq:e29upper}
    	\bbP[Z_n \leq \delta n] \leq \frac{\bbE\off{e^{\lambda Z_n}}}{e^{\lambda \delta n}} = \offf {\exp \offf{\log \of{1 - \alpha + \alpha e^{\lambda})}- \lambda \delta }}^n,\quad \text{for all } \lambda \in \bbR,
    \end{equation}
    where the second equality follows from the moment generating function of $Z_n$
    \[\bbE\off{e^{\lambda Z_n}} = \of{1 - \alpha + \alpha e^{\lambda}}^n. \]
    The upper bound~\eqref{eq:e29upper} achieves the minimum when $\lambda = \log {\frac{\delta(1-\alpha)}{\alpha(1 - \delta)}}$ by the first-order condition. Plugging the minimizer into equation~\eqref{eq:e29upper}, we obtain the result
    \[ \bbP\off{Z_n \leq \delta n}  \leq e^{-n D\of{\delta \Vert \alpha}}, \quad \text{for all }\delta \in (0,\alpha).  \]

    \vspace{0.2cm}
    Next, we show that equation~\eqref{eq:lembinupper} is strictly better than Hoeffding bound for all $\delta \in (0,\alpha)$.

    \vspace{0.2cm}
    Hoeffding bound for random variable $Z_n$ is 
    \begin{equation}\label{eq:e29hoeff}
    	\bbP\off{Z_n \leq \delta n} \leq e^{ {-2n(\delta - \alpha)^2}}, \quad \text{for all }\delta \in (0,\alpha).
    \end{equation}
    Note that parameter $\alpha$ is fixed. Consider the function $g(\delta) = 2(\delta - \alpha)^2 - D\of{\delta \Vert \alpha}$ for $\delta \in (0,\alpha)$. The first-order derivative and  second-order derivative of $g$ are
    \[ g'(\delta) = 4(\delta - \alpha) - \log \frac{\delta}{\alpha} + \log \frac{1-\delta}{1-\alpha} \quad \text{and} \quad g''(\delta) = 4 - \frac{1}{\delta} - \frac{1}{1-\delta}. \] 
    Note that $ g(\alpha) = g'(\alpha) = 0$, and $g''(\delta) < 0$ for all $0 < \delta < 1/2$. We have $g(\delta) < 0$, for all $\delta \in (0,\alpha)$. Hence, the upper bound~\eqref{eq:lembinupper} is strictly better than Hoeffding bound~\eqref{eq:e29hoeff}.

@@

\begin{rmk}
	The upper bound on a binomial tail may be a function of the KL divergence between Bernoulli distributions with distinct parameters. The bound using KL divergence is strictly tighter than the Hoeffding bound. The underperformance of Hoeffding bound may attribute to the utilization of the sub-Gaussian parameter $\sigma^2$. The sub-Gaussian parameter $\sigma^2$ is not necessarily the optimal choice to describe the tail performance of a random variable with good properties. 
\end{rmk}
\end{document}

>>>>

\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}

@@

	To evaluate $I_1$, we spilt the integral to avoid the minimization. Solving $1 = C \exp \of{-\frac{t^2}{4 \nu^2}} $, the minimization term becomes
	\begin{align}
		\min \offf{1, C \exp \of{-\frac{t^2}{4 \nu^2}}} = \begin{cases}
			\quad 1 & \text{when } t < 2\nu \sqrt{\log C},\\
			\quad C \exp \of{-\frac{t^2}{4 \nu^2}} & \text{when } t \geq 2\nu \sqrt{\log C}.
		\end{cases}
	\end{align}
	Therefore, we evaluate $I_1$ as follows,
	\begin{align}
		I_1 &= \int_{0}^{ 2\nu \sqrt{\log C}} 1 dt + \int_{2\nu \sqrt{\log C}}^{+\infty} C \exp \of{-\frac{t^2}{4 \nu^2}} dt\\
		(\text{let}\quad y = \frac{t}{2\nu} - \sqrt{\log C})\quad &=  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2 - 2y \sqrt{\log C}} dy\\
		&\leq  2\nu \sqrt{\log C} + 2\nu\int_{0}^{+\infty} \exp \of{-y^2} dy.
	\end{align}
	By Gaussian integral $\int_{0}^{+\infty} e^{-x^2} dx = \frac{\sqrt{\pi}}{2}$, we obtain $I_1 \leq 2\nu (\sqrt{\pi} + \sqrt{\log C})$. Similarly, we evaluate $I_2$ as follows 
	\begin{align}
		I_2 &= \int_{0}^{4B \log C} 1 dt + \int_{4B \log C}^{+\infty} C \exp \of{-\frac{t}{4B}} dt\\
		 &= 4B (\log C + 1).
	\end{align}
	Hence, we obtain the expectation of $Z$,
	\begin{equation}
		\bbE[Z] = I_1 + I_2 \leq  2\nu (\sqrt{\pi} + \sqrt{\log C}) + 4B (\log C + 1).
	\end{equation}

@@

\begin{rmk}
	For a nonnegative random variables satisfying the Bernstein-type inequality \eqref{eq:berncond1}, the expectation of the variable is upper bounded by a function of the parameter $(\nu,B,C)$. Particularly, for a zero-mean random variable $X$ satisfying the Bernstein condition with parameter $b$, $|X|$ satisfies the inequality \eqref{eq:berncond1} with parameters $(\sigma,b,2)$, where $\sigma^2 = \var(X)$.  Then, the expectation of the absolute variable $|X|$ is upper bounded by a function of $(\sigma^2, b)$. 
\end{rmk}

\subsection{Exercise 2.9}
\begin{lem}[Sharp upper bounds on binomial tails]
	Let $\{X_i\}_{i=1}^n$ be a sequence of i.i.d.\ Bernoulli variables with parameter $\alpha \in (0,1/2]$. Consider the binomial random variable $Z_n = \sum_{i=1}^n X_i$. The tail probability of $Z_n$ is upper bounded as
	\begin{equation}\label{eq:lembinupper}
		\bbP\off{Z_n \leq \delta n} \leq e^{-n D\of{\delta \Vert \alpha}}, \quad \text{for all }\delta \in (0,\alpha),
	\end{equation}
	where the quantity 
	\begin{equation}\label{eq:kldiv}
		D(\delta \Vert \alpha) \colonequals \delta \log \frac{\delta}{\alpha} + (1-\delta) \log \frac{1-\delta}{1-\alpha}
	\end{equation}
	is the KL divergence between the Bernoulli distributions with parameters $\delta$ and $\alpha$, respectively.

@@

\begin{proof}
    Applying the Chernoff inequality to the random variable $Z_n$, we have
    \begin{equation}\label{eq:e29upper}
    	\bbP[Z_n \leq \delta n] \leq \frac{\bbE\off{e^{\lambda Z_n}}}{e^{\lambda \delta n}} = \offf {\exp \offf{\log \of{1 - \alpha + \alpha e^{\lambda})}- \lambda \delta }}^n,\quad \text{for all } \lambda \in \bbR,
    \end{equation}
    where the second equality follows from the moment generating function of $Z_n$,
    \[\bbE\off{e^{\lambda Z_n}} = \of{1 - \alpha + \alpha e^{\lambda}}^n. \]
    The upper bound~\eqref{eq:e29upper} achieves the minimum when $\lambda = \log {\frac{\delta(1-\alpha)}{\alpha(1 - \delta)}}$ by the first-order condition. Plugging the minimizer into equation~\eqref{eq:e29upper}, we obtain the result
    \[ \bbP\off{Z_n \leq \delta n}  \leq e^{-n D\of{\delta \Vert \alpha}}, \quad \text{for all }\delta \in (0,\alpha).  \]

    \vspace{0.2cm}
    Next, we show that equation~\eqref{eq:lembinupper} is strictly tighter than the Hoeffding bound of $Z_n$, for all $\delta \in (0,\alpha)$.

    \vspace{0.2cm}
    Hoeffding bound of random variable $Z_n$ is 
    \begin{equation}\label{eq:e29hoeff}
    	\bbP\off{Z_n \leq \delta n} \leq e^{ {-2n(\delta - \alpha)^2}}, \quad \text{for all }\delta \in (0,\alpha).
    \end{equation}
    Note that parameter $\alpha$ is fixed. Consider the function $g(\delta) = 2(\delta - \alpha)^2 - D\of{\delta \Vert \alpha}$ for $\delta \in (0,\alpha)$. The first-order derivative and  second-order derivative of $g$ are
    \[ g'(\delta) = 4(\delta - \alpha) - \log \frac{\delta}{\alpha} + \log \frac{1-\delta}{1-\alpha} \quad \text{and} \quad g''(\delta) = 4 - \frac{1}{\delta} - \frac{1}{1-\delta}. \] 
    Note that $ g(\alpha) = g'(\alpha) = 0$, and $g''(\delta) < 0$ for all $0 < \delta < 1/2$. Then, we have $g(\delta) < 0$, for all $\delta \in (0,\alpha)$. Hence, the upper bound~\eqref{eq:lembinupper} is strictly tighter than Hoeffding bound~\eqref{eq:e29hoeff}.

@@

\begin{rmk}
	The upper bound on a binomial tail is a function of the KL divergence between Bernoulli distributions with distinct parameters. The bound using KL divergence is strictly tighter than the Hoeffding bound. The underperformance of Hoeffding bound may attribute to the utilization of the sub-Gaussian parameter $\sigma^2$. The sub-Gaussian parameter $\sigma^2$ is not necessarily the optimal choice to describe the tail performance of a random variable with good properties. 
\end{rmk}

\subsection{Exercise 2.10}
\begin{lem}[Lower bounds on binomial tails]
	Let $\{X_i\}_{i=1}^n$ be a sequence of i.i.d.\ Bernoulli variables with parameter $\alpha \in (0,1/2]$. Consider the binomial random variable $Z_n \colonequals \sum_{i=1}^n X_i$. For some fixed $\delta \in (0,\alpha)$, let $m = \floor{n\delta}$; i.e., $m$ is the largest integer less or equal to $n \delta$. Let $ \tilde \delta = {m}/{n}$. We have
	\begin{equation}\label{eq:e2101}
		\frac{1}{n} \log \bbP \off{Z_n \leq n\delta} \geq \frac{1}{n} \log \begin{pmatrix}
			n \\ m	\end{pmatrix} + \tilde \delta \log \alpha + (1 - \tilde \delta ) \log (1-\alpha).
	\end{equation}

	Further, the binomial coefficient satisfies
	\begin{equation}\label{eq:e2102}
		\frac{1}{n} \log \begin{pmatrix}
			n \\ m	\end{pmatrix} \geq \phi(\tilde \delta ) - \frac{\log (n+1)}{n}.
	\end{equation}

	Consequently, the lower tail of binomial variable $Z_n$ satisfies
	\begin{equation}\label{eq:e2103}
		\bbP[Z_n \leq n \delta]  \geq \frac{1}{n+1} e^{-nD(\delta \Vert \alpha)},
	\end{equation}
	where $D(\delta \Vert \alpha)$ is the KL divergence defined in equation~\eqref{eq:kldiv}.
\end{lem}

\begin{proof}
	First, we prove equation~\eqref{eq:e2101}.

	\vspace{0.2cm}
	Since $Z_n$ is a binomial variable with size $n$ and probability $\alpha$, we have
	\begin{align}
		\bbP[Z_n \leq n\delta ] &= \sum_{k = 1}^m \begin{pmatrix}
			n \\ k
		\end{pmatrix} \alpha^k (1-\alpha)^{n - k}\\
		& \geq \begin{pmatrix}
			n \\ m
		\end{pmatrix} \alpha^m (1-\alpha)^{n - m}. \label{eq:e210pro1}
	\end{align}
	Taking the log and dividing by $n$ on the both sides of the inequality~\eqref{eq:e210pro1} , we obtain
	\[ \frac{1}{n} \log \bbP[Z_n \leq n\delta ] \geq \frac{1}{n} \log \begin{pmatrix}
			n \\ m		\end{pmatrix} + \tilde \delta \log \alpha + (1 - \tilde \delta ) \log (1-\alpha). \]

	\vspace{0.2cm}
	Next, we prove equation~\eqref{eq:e2102}.

	\vspace{0.2cm}
	Consider a binomial variable $Y \sim Bin(n,\tilde \delta)$. For all $k = 0,1,...,(n-1)$, the ratio between $\bbP[Y = k]$ and $\bbP[Y = k+1]$ is
	\begin{equation}
		\frac{\bbP[Y = k+1]}{\bbP[Y = k]} = \frac{\begin{pmatrix}
			n \\ k+1
		\end{pmatrix} \tilde \delta^{k+1} (1-\tilde \delta)^{n - k -1} }{\begin{pmatrix}
			n \\ k
		\end{pmatrix} \tilde \delta^{k} (1-\tilde \delta)^{n - k }} = \frac{(n-k) \tilde \delta}{(k+1) (1- \tilde \delta)  }.
	\end{equation}
	To let the ratio $	\frac{\bbP[Y = k+1]}{\bbP[Y = k]} \geq 1$, we need $(k+1)\leq (n+1) \tilde \delta $. Thus, the probability $\bbP[Y = l]$ achieves the maximum when $l = n \tilde \delta$. Consequently, we have
	\begin{equation}\label{eq:e210pro2}
		(n+1) \bbP[Y = n \tilde \delta] \geq 1 \quad \Leftrightarrow \quad  \bbP[Y = n \tilde \delta] \geq \frac{1}{n+1}.
	\end{equation}
	Taking the log and dividing by $n$ on the both sides of the inequality~\eqref{eq:e210pro2}, we obtain
	\begin{equation}
		 \frac{1}{n} \log \begin{pmatrix}
			n \\ m		\end{pmatrix} \geq \tilde \delta \log(\alpha) - (1 - \tilde \delta)\log(1 - \alpha) - \frac{\log(n+1)}{n} = \phi(\tilde \delta)  - \frac{\log(n+1)}{n}.
	\end{equation}

	\vspace{0.2cm}
	Last, we prove the lower bound equation~\eqref{eq:e2103}.

	\vspace{0.2cm}
	Plugging equation~\eqref{eq:e2102} into equation~\eqref{eq:e2101}, we have
	\begin{equation}\label{eq:e210pro3}
		\frac{1}{n} \log \bbP \off{Z_n \leq n\delta} \geq \phi(\tilde \delta) + \tilde \delta \log \alpha + (1 - \tilde \delta ) \log (1-\alpha) - \frac{\log(n+1)}{n} \geq -D(\delta \Vert \alpha ) - \frac{\log(n+1)}{n}.
	\end{equation}
	Multiplying $n$ and exponentiating on the both sides of the inequality~\eqref{eq:e210pro3}, we obtain the result
	\begin{equation}
		\bbP \off{Z_n \leq n\delta} \geq \frac{1}{n+1} e^{-nD(\delta \Vert \alpha)}.
	\end{equation}
\end{proof}

\begin{rmk}
	The lower bound on a binomial tail is also a function of the KL divergence between Bernoulli distributions with distinct parameters. Combining the upper bound \eqref{eq:lembinupper} and lower bound \eqref{eq:e2103}, we conclude that the binomial tail is (upper and lower) bounded by the functions of  KL divergence between Bernoulli distributions.
\end{rmk}

\subsection{Exercise 2.11}
\begin{lem}[Gaussian maxima]
Let $\{X_i\}_{i=1}^n$ be a sequence of i.i.d. normal random variables following $N(0,\sigma^2)$. Consider the random variable $Z_n \colonequals \max_{i\in[n]} |X_i|$. Since the tail bound $\bbP[U \geq t] \leq \sqrt{\frac{2}{\pi}} \frac{1}{t} e^{-t^2 /2}$ holds for all standard normal random variable $U$, the expectation of $Z_n$ is upper bounded as follows
\begin{equation}\label{eq:e2111}
	\bbE[Z_n] \leq \sqrt{2 \sigma^2 \log n} + \frac{4\sigma}{\sqrt{2 \log n}}, \quad \text{for all } n\geq 2.
\end{equation}
\iffalse
The lower bound of $\bbE[Z_n] $ is
\begin{equation}\label{eq:e2112}
	\bbE[Z_n] \geq (1- 1/e) \sqrt{2\sigma^2 \log n} , \quad \text{for all }n \geq 5.
\end{equation}
\fi

\end{lem}

\begin{proof}
	%First, we prove the equation~\eqref{eq:e2111}.

	\vspace{0.2cm}
	Consider the tail probability of $Z_n$. We have
	\begin{equation}
		\bbP[Z_n \geq t] = \bbP[\max_{i \in [n]} |X_i| \geq t] = 1 - \bbP[\max_{i \in [n]} |X_i| < t] = 1 - (1- \bbP[|X_1| \geq t])^n,
	\end{equation}
	where the last equality follows from the independence of $\{X_i\}_{i=1}^n$. By Bernoulli's inequality, we have
	\begin{equation}
		(1- \bbP[|X_1| \geq t])^n \geq 1 - n \bbP[|X_1| \geq t], \quad \text{for all } t >0.
	\end{equation} 
	Therefore, the expectation of $Z_n$ follows
	\begin{equation}\label{eq:e211pro1}
		\bbE[Z_n] = \int_{0}^{+\infty} \bbP[Z_n \geq t] dt \leq  c +  \int_{c}^{+\infty} n\bbP[|X_1| \geq t] dt, \quad \text{for all } c > 0.
	\end{equation}
	 Since the tail bound $\bbP[U \geq t] \leq \sqrt{\frac{2}{\pi}} \frac{1}{t} e^{-t^2 /2}$ holds for all standard normal random variable $U$, the tail bound of $|X_1|$ satisfies
	 \begin{equation}\label{eq:e211proupper}
	 	\bbP[|X_1| \geq t] \leq 2 \sqrt{\frac{2}{\pi}} \frac{\sigma}{t} e^{-\frac{t^2}{2\sigma^2}}.
	 \end{equation}
	 Hence, the last integral in equation~\eqref{eq:e211pro1} follows
	 \begin{align}
	 	 \int_{c}^{+\infty} n\bbP[|X_1| \geq t] dt &\leq \frac{2n \sigma}{c} \sqrt{\frac{2}{\pi}} \int_{c}^{+\infty} e^{-\frac{t^2}{2\sigma^2} }dt \\ (\text{let } u = \frac{ t}{\sigma}) \quad  
	 	 %& = \frac{2n \sigma^2}{c} \sqrt{\frac{2}{\pi}}   \int_{\frac{c}{\sigma}}^{+\infty} e^{-\frac{u^2}{2}} dt \\
	 	 & \leq \frac{2n \sigma^2}{c} \sqrt{\frac{2}{\pi}}   \int_{\frac{c}{\sigma}}^{+\infty} u e^{-\frac{u^2}{2}} dt \\
	 	 & = \frac{2n \sigma^2}{c} \sqrt{\frac{2}{\pi}}  e^{-\frac{c^2}{2\sigma^2}}. \label{eq:e211pro2}
	 \end{align}
	 For all $n \geq 2$, let $c = \sqrt{2\sigma^2 \log n}$ and plug the $c$ into equations~\eqref{eq:e211pro1} and \eqref{eq:e211pro2}. Then, we obtain the result
	 \begin{equation}
	 	\bbE[Z_n] \leq \sqrt{2\sigma^2 \log n} + \frac{2\sigma}{\sqrt{2\log n}}   \sqrt{\frac{2}{\pi}}  \leq \sqrt{2\sigma^2 \log n} + \frac{4\sigma}{\sqrt{2\log n}}, \quad \text{for all }n \geq 2.
	 \end{equation}

	 \iffalse
	 \vspace{0.2cm}
	 Next, we prove the equation~\eqref{eq:e2112}.

	 \vspace{0.2cm}

	 By Markov's inequality, the expectation of $Z_n$ satisfies
	 \begin{equation}
	 	\bbE[Z_n] \geq t \bbP[Z_n \geq t],\quad \text{for all }t \geq 0.
	 \end{equation}
	 By the independence of $\{X_i\}_{i=1}^n$, the tail probability of $Z_n$ is
	 \begin{equation}
	 	\bbP[Z_n \geq t] = 1 - (1 - \bbP(|X_1| \geq t))^n.
	 \end{equation}

	 Let $t = \sqrt{2\sigma^2 \log n}$. By equation~\eqref{eq:e211proupper}, the tail probability of $X_1$ satisfies
	 \begin{equation}
	 	\bbP(|X_1| \geq \sqrt{2\sigma^2 \log n}) \leq \frac{2}{\sqrt{ \pi \log n}} \frac{1}{n} \leq \frac{1}{n},\quad \text{for all } n \geq 5.
	 \end{equation}	 
	 Hence, the tail probability of $Z_n$ with $t = \sqrt{2\sigma^2 \log n} $satisfies
	 \begin{equation}
	 	\bbP[Z_n \geq t] \geq 1 - 
	 \end{equation}
	 \fi
\end{proof}

\end{document}
